---
title: "ggforce: Visual Guide"
author: "Thomas Lin Pedersen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
This document serves as the main overview of the ggforce package, and will try 
to explain the hows and whys of the different extension along with clear visual
examples. It will try to link back to relevant academic articles describing the
different visualization types in more detail - both for the benefit of the 
reader but also to give credit to the people who thought long and hard about how
to best present your data.

We'll start by loading ggforce and then proceed to all the beautiful graphics...

```{r, echo=TRUE, eval=TRUE}
library(ggforce)
```

# Layers
This section shows the extensions to ggplot2's geoms and stats. It rarely makes 
sense to talk about one and not the other, so they are grouped together here. 
Often the focus will be on the geoms, unless a new stat does not have an 
accompanying geom, in which case the stat will be discussed along with which
geoms it should be used with.

## Arcs
Arcs are segments of a circle and defined by a centre point, a radius and a 
start and end angle. In ggforce arcs come in two flavors: arc and arc_bar, where
the former draws an arc with a single line and the latter draws it as a polygon
that can have a fill and outline. A wedge is a special case of arc_bar where the 
innermost radius is 0. The most well known use of arcs in plotting is with the
much loathed pie chart (and its cousin the donut chart). The reason for all the
hatred against pie charts are just and related to the fact that humans are much 
better at comparing heights than angles. Because of this a bar chart will always
communicate your data better than a pie chart. Donut charts are a little better
as the hole in the middle forces the eye to compare arc spans rather than 
angles, but it is still better to use a bar chart. Arcs, being a fundamental 
visual element, can be used for other things though, such as sunburst plots or
annotating radial visualizations.

As pie charts are most well known, we'll start by upsetting all visualization
expert and produce one:

```{r, eval=TRUE, echo=TRUE, fig.align='center'}
# We'll start by defining some dummy data
pie <- data.frame(
    state = c('eaten', 'eaten but said you didn\'t', 'cat took it', 
              'for tonight', 'will decompose slowly'),
    focus = c(0.2, 0, 0, 0, 0),
    start = c(0, 1, 2, 3, 4),
    end = c(1, 2, 3, 4, 2*pi),
    amount = c(4,3, 1, 1.5, 6),
    stringsAsFactors = FALSE
)

p <- ggplot() + theme_no_axes() + coord_fixed()

# For low level control you define the start and end angles yourself
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, start = start, end = end, 
                     fill = state),
                 data = pie)

# But often you'll have values associated with each wedge. Use stat_pie then
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, 
                     fill = state),
                 data = pie, stat = 'pie')

# The wedges can be exploded away from the centre using the explode aesthetic
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, 
                     fill = state, explode = focus),
                 data = pie, stat = 'pie')

# And a donut can be made by setting r0 to something > 0
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0.8, r = 1, amount = amount, 
                     fill = state, explode = focus),
                 data = pie, stat = 'pie')

```

While the above produces some of the most hated plot types in the world it does
showcase the use of arcs in plotting. Other examples will use geom_arc and 
geom_arc_bar to produce more powerful plot types.

## Edge bundles
When using edges to show connections you often end up with a lot of overploting,
effectively concealing the underlying structure of the connections. To combat 
this Danny Holten invented a technique called 
[*Hierarchical Edge Bundles*](http://www.aviz.fr/wiki/uploads/Teaching2014/bundles_infovis.pdf)
applicable to situations where the connected nodes has an inherent hierarchy.
The technique works by letting the connections follow the route defined by the
hierarchy thus clearly showing main trends in the connections with regards to
the node structure. While the technique was first proposed for hierarchical 
data, it can just as well be applied to other situations where paths between 
points can be mapped onto a common grid.

In ggforce hierarchical edge bundles are implemented with the 
geom_edge_bundle/stat_edge_bundle functions, that takes the control points for
each connection and creates the bundles. As there is a lot going on prior to
plotting (defining the hierarchical layout, getting the paths for each 
connection etc.) the example is quite long - once you have the control points 
though, it is as simple as using geom_path

```{r, echo=TRUE, eval=TRUE, fig.align='center'}
# Get the flare class hierarchy
data('flare', package = 'ggforce')

# Convert to an edgelist of dependencies
flareH <- strsplit(flare$name, '.', fixed = TRUE)
edges <- data.frame(
    from = character(), 
    to = character(), 
    stringsAsFactors = FALSE
)

while (TRUE) {
    flareH <- flareH[lengths(flareH) > 1]
    if (length(flareH) == 0) break
    connect <- unique(lapply(flareH, `[`, i = 1:2))
    edges <- rbind(edges, data.frame(
        from = sapply(connect, `[`, i = 1), 
        to = sapply(connect, `[`, i = 2), 
        stringsAsFactors = FALSE
        )
    )
    flareH <- lapply(flareH, `[`, i = -1)
}

# Create graph and define layout function
gr <- igraph::graph_from_edgelist(as.matrix(edges))
layout_cluster <- function(gr) {
    nodes <- data.frame(
        x = rep(NA_real_, igraph::gorder(gr)), 
        y = rep(NA_real_, igraph::gorder(gr)), 
        leaf = igraph::degree(gr, mode = 'out') == 0, 
        stringsAsFactors = FALSE
    )
    startnode <- which(igraph::degree(gr, mode = 'in') == 0)
    recurse_layout <- function(gr, node, layout) {
        children <- as.numeric(igraph::neighbors(gr, node, 'out'))
        if (length(children) == 0) {
            x <- if (all(is.na(layout$x[layout$leaf]))) {
                1
            } else {
                max(layout$x[layout$leaf], na.rm = TRUE) + 1
            }
            layout$x[node] <- x
            layout$y[node] <- 0
            layout
        } else {
            childrenMissing <- children[is.na(layout$x[children])]
            for (i in childrenMissing) {
                layout <- recurse_layout(gr, i, layout)
            }
            layout$x[node] <- mean(layout$x[children])
            layout$y[node] <- max(layout$y[children]) + 1
            layout
        }
    }
    recurse_layout(gr, startnode, nodes)
}

# Calculate and refine the layout, we'll use a radial layout
layout <- layout_cluster(gr)
layout$angle[layout$leaf] <- layout[layout$leaf,'x']/221 * -360 - 270 %% 360
layout$label <- rownames(layout)
radial <- radial_trans(rev(range(layout$y)), range(layout$x) + c(0, 1))
layoutR <- radial$transform(layout$y, layout$x)
layout$x <- layoutR$x
layout$y <- layoutR$y
layout$angle[layout$leaf & layout$x < 0] <- 
    layout$angle[layout$leaf & layout$x < 0] + 180

# Get the paths for all dependencies
paths <- do.call(rbind, lapply(seq_len(nrow(flare)), function(i) {
    imports <- sapply(
        strsplit(flare$imports[[i]], '.', fixed = TRUE), 
        tail, 
        n = 1
    )
    do.call(rbind, lapply(imports, function(j) {
        path <- igraph::all_simple_paths(gr, flare$shortName[i], j, mode = 'all')
        path <- as.numeric(path[[which.min(lengths(path))]])
        data.frame(
            x = layout[path, 1], 
            y = layout[path, 2], 
            group = paste0(flare$shortName[i], '->', j)
        )
    }))
}))

# Now for the plotting
ggplot() + 
    coord_fixed() + theme_void() + # Setup plot
    # Add edge bundles
    geom_edge_bundle(aes(x = x, y = y, group = group, colour = ..index..), 
                     data = paths, alpha = 0.2, tension = 0.9) + 
    # Add labels to the leafs
    geom_text(aes(x = x, y = y, label = label, angle = angle, vjust = 0.5, 
                  hjust = ifelse(x < 0, 1, 0)), 
              data = layout[layout$leaf,], size = 2.5) + 
    # Format scales and make room for the labels
    scale_color_gradient(low='goldenrod', high='firebrick', guide='none') + 
    expand_limits(x=c(-1.3, 1.3), y=c(-1.3, 1.3))
```

Contrast this to not using edge bundles (equivalent to setting tension = 0) and
the benefits becomes obvious.

```{r, echo=TRUE, eval=TRUE, fig.align='center'}
ggplot() + 
    coord_fixed() + theme_void() + 
    geom_edge_bundle(aes(x = x, y = y, group = group, colour = ..index..), 
                     data = paths, alpha = 0.2, tension = 0) + 
    geom_text(aes(x = x, y = y, label = label, angle = angle, vjust = 0.5, 
                  hjust = ifelse(x < 0, 1, 0)), 
              data = layout[layout$leaf,], size = 2.5) + 
    scale_color_gradient(low='goldenrod', high='firebrick', guide='none') + 
    expand_limits(x=c(-1.3, 1.3), y=c(-1.3, 1.3))
```

While hierarchical edge bundles are often used with radial layout, they are not
confined to this. The radial layout has the benefit of minimizing the edge 
overlaps though, as can be seen when we plot it without any radial 
transformation:

```{r, echo=TRUE, eval=TRUE, fig.align='center'}
layout <- layout_cluster(gr)
layout$label <- rownames(layout)

paths <- do.call(rbind, lapply(seq_len(nrow(flare)), function(i) {
    imports <- sapply(
        strsplit(flare$imports[[i]], '.', fixed = TRUE), 
        tail, 
        n = 1
    )
    do.call(rbind, lapply(imports, function(j) {
        path <- igraph::all_simple_paths(gr, flare$shortName[i], j, mode = 'all')
        path <- as.numeric(path[[which.min(lengths(path))]])
        data.frame(
            x = layout[path, 1], 
            y = layout[path, 2], 
            group = paste0(flare$shortName[i], '->', j)
        )
    }))
}))

ggplot() + 
    theme_void() + 
    geom_edge_bundle(aes(x = x, y = y, group = group, colour = ..index..), 
                     data = paths, alpha = 0.2, tension = 0.9) + 
    geom_text(aes(x = x, y = y, label = label, angle = 45, vjust = 0.5, 
                  hjust = 1), 
              data = layout[layout$leaf,], size = 1.5) + 
    scale_color_gradient(low='goldenrod', high='firebrick', guide='none')
```

# Transformations
Transformations are not really a part of ggplot2, but rather the scales package.
Nevertheless it is an integral part of working with ggplot2 through its use in
manipulating scales. ggforce expands the use of transformations to also include
coordinate transformations.

## Univariate transformations
This section describes the new transformations offered by ggforce for 
manipulating scales. In general the scales package has you well covered but 
there are some missing pieces:

### Power transformations
TBD

### Reversing transformations
TBD

## Coordinate transformations
Coordinate transformation takes coordinates and does something to them. It can
be simple rotations, shearing and reflections as you know from different image
processing applications, or translating between different ways of representing
data, e.g. radial to cartesian transformations. These types of transformations
are closely linked to applying different coordinate systems to your plot, e.g.
using coord polar, but can be applied to your data upfront instead of on the
whole plot.

### Radial transformations
TBD
