---
title: "ggforce: Visual Guide"
author: "Thomas Lin Pedersen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
This document serves as the main overview of the ggforce package, and will try 
to explain the hows and whys of the different extension along with clear visual
examples. It will try to link back to relevant academic articles describing the
different visualization types in more detail - both for the benefit of the 
reader but also to give credit to the people who thought long and hard about how
to best present your data.

We'll start by loading ggforce and then proceed to all the beautiful graphics...

```{r, echo=TRUE, eval=TRUE}
library(ggforce)
```

## Geom versions
TBD

# Layers
This section shows the extensions to ggplot2's geoms and stats. It rarely makes 
sense to talk about one and not the other, so they are grouped together here. 
Often the focus will be on the geoms, unless a new stat does not have an 
accompanying geom, in which case the stat will be discussed along with which
geoms it should be used with.

## Arcs
Arcs are segments of a circle and defined by a centre point, a radius and a 
start and end angle. In ggforce arcs come in two flavors: arc and arc_bar, where
the former draws an arc with a single line and the latter draws it as a polygon
that can have a fill and outline. A wedge is a special case of arc_bar where the 
innermost radius is 0. The most well known use of arcs in plotting is with the
much loathed pie chart (and its cousin the donut chart). The reason for all the
hatred against pie charts are just and related to the fact that humans are much 
better at comparing heights than angles. Because of this a bar chart will always
communicate your data better than a pie chart. Donut charts are a little better
as the hole in the middle forces the eye to compare arc spans rather than 
angles, but it is still better to use a bar chart. Arcs, being a fundamental 
visual element, can be used for other things though, such as sunburst plots or
annotating radial visualizations.

As pie charts are most well known, we'll start by upsetting all visualization
expert and produce one:

```{r, eval=TRUE, echo=TRUE, fig.align='center'}
# We'll start by defining some dummy data
pie <- data.frame(
    state = c('eaten', 'eaten but said you didn\'t', 'cat took it', 
              'for tonight', 'will decompose slowly'),
    focus = c(0.2, 0, 0, 0, 0),
    start = c(0, 1, 2, 3, 4),
    end = c(1, 2, 3, 4, 2*pi),
    amount = c(4,3, 1, 1.5, 6),
    stringsAsFactors = FALSE
)

p <- ggplot() + theme_no_axes() + coord_fixed()

# For low level control you define the start and end angles yourself
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, start = start, end = end, 
                     fill = state),
                 data = pie)

# But often you'll have values associated with each wedge. Use stat_pie then
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, 
                     fill = state),
                 data = pie, stat = 'pie')

# The wedges can be exploded away from the centre using the explode aesthetic
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, 
                     fill = state, explode = focus),
                 data = pie, stat = 'pie')

# And a donut can be made by setting r0 to something > 0
p + geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0.8, r = 1, amount = amount, 
                     fill = state, explode = focus),
                 data = pie, stat = 'pie')
```

While the above produces some of the most hated plot types in the world it does
showcase the use of arcs in plotting. Other examples will use geom_arc and 
geom_arc_bar to produce more powerful plot types. 

Using arc is just like arc_bar except that it does not take an r0 argument and
does not have any fill.

## Circles
Standard ggplot2 generally has you covered when it comes to drawing circles
through the point geom, it does not make it possible to draw circles where the
radius of the circles are related to the coordinate system. The geom_circle from
ggforce are precisely for that. It generates a polygon resembling a circle based 
on a center point and a radius, making the radius directly readable from the 
axes. The geom are mainly intended to make it possible to draw circles with fine 
grained control, but will often not have any utility in itself. An exception 
would be in plotting trees as enclosure diagrams using circles. Here it will be
necessary to have fine control over radius.

```{r, echo=TRUE, eval=TRUE, fig.align='center'}
# Here are some data describing some circles
circles <- data.frame(
    x0 = rep(1:3, 2),
    y0 =  rep(1:2, each=3),
    r = seq(0.1, 1, length.out = 6)
)
ggplot() + geom_circle(aes(x0=x0, y0=y0, r=r, fill=r), data=circles)

# As it is related to the coordinate system, coord_fixed() is needed to ensure
# true circularity
ggplot() + geom_circle(aes(x0=x0, y0=y0, r=r, fill=r), data=circles) +
    coord_fixed()

# Use n to set the smoothness of the circle
ggplot() + geom_circle(aes(x0=x0, y0=y0, r=r, fill=r), data=circles, n=10) +
    coord_fixed()
```

## Links
TBD

## Beziers
TBD

## B-splines
TBD

# Transformations
Transformations are not really a part of ggplot2, but rather the scales package.
Nevertheless it is an integral part of working with ggplot2 through its use in
manipulating scales. ggforce expands the use of transformations to also include
coordinate transformations.

## Univariate transformations
This section describes the new transformations offered by ggforce for 
manipulating scales. In general the scales package has you well covered but 
there are some missing pieces:

### Power transformations
TBD

### Reversing transformations
TBD

## Coordinate transformations
Coordinate transformation takes coordinates and does something to them. It can
be simple rotations, shearing and reflections as you know from different image
processing applications, or translating between different ways of representing
data, e.g. radial to cartesian transformations. These types of transformations
are closely linked to applying different coordinate systems to your plot, e.g.
using coord polar, but can be applied to your data upfront instead of on the
whole plot.

### Radial transformations
TBD
